---
title: "A look into Kamala Harris' polls in the upcoming 2024 US Presidential Election "
subtitle: ""
author: 
  - Veyasan Ragulan
thanks: "Code and data are available at: https://github.com/Veyasan1/2024USElectionModel"
date: today
date-format: long
abstract: "The 2024 US election has been seen as a close race between the Democratic nominess Kamala Harris, and the Republican Nominee Donald Trump, with each candidate holding down key states and matching each other in battleground states such as Pennsylvania and Georgia. This paper looks at the polling as of late October 2024 to see how Kamala Harris fares as the election draws near."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

library(rstanarm)

bayesian_model <-
  readRDS(file = here::here("models/electionmodel.rds"))

spline_model <- 
  readRDS(file = here::here("models/splinemodel.rds"))
```

# Introduction

In this paper we look at polling data from 538 on the 2024 US Presidential election, specifically on Demoncratic Candidate Kamala Harris. We will construct a model that will be used to make a prediction on whether Harris wins the US Presidency

Section 2 will outline the source of this data. Section 3 covers the model and it's parameters. Section 4 is where discussion will be made about the models predictions and how realistically they line up with current affairs. Finally, section 4 discusses any weakness and limitations that can be considered for another report.

This report was written with the assistance of R @citeR. Additionally, @tidyverse, and @rstanarm were used to clean the data, write models, and analyse results. The structure of this report heeds Rohan Alexander's example @tellingstories.

# Data {#sec-data}

The data used in this model comes from fivethirtyeight.com, a polling site known for their in-depth analysis and data on US elections @fivethirtyeight. Here, we have used the 2024 general election data, originally accessed on 10/21/2024. The data contains information from several polls taken across the United States, including the operator, sponsors, methodology, a numeric grade to quanitfy the quality of the poll, and when and where the poll was taken. From this dataset, polls predicting a Harris win were considered, to construct the predictive model.

# Model

The model contains 4 parameters. They are pollster, state, sponsors, and methodology.

Pollster is the operator of the observed poll. Pollster provides a relatively easy way to identify trends in certain polls, depending on who is conducting them.

State tells us where the poll was conducted. Across the United States, ther are states that traditionally vote Democrat (New York, California), some that traditionally vote Republican (Texas, Florida), and 'swing states' (Michigan, Pennsylvania). Swing states in particular are the most interesting states to watch as votes from these states tend to win a candiate the US presidency.Knowing where the polls have been taken also lets us understand if their choice was one bound by the state's culture.

Sponsors are organizations that fund or otherwise assist the pollster in operating the poll. Sponsors carry the risk of bias with them, potentially influencing the pollster, and thus the polls to skew in a direction the sponsor favors preferably. Knowing who these sponsors are potemntially lets us make informed predictions on the outcome of polls.

Methodology is the way the poll was conducted. Sometimes the composition of the poll influences peoples votes, so this is an interesting parameter to track. This also lets us see if certain methods yield better or worse predictions.

# Results

```{r}
#| echo: false
#| warning: false
#| message: false

# Posterior predictive checks
pp_check(bayesian_model)

# Summarize the model
summary(bayesian_model)

# Plot random effects
plot(bayesian_model, pars = "(Intercept)", prob = 0.95)

```

```{r}
#| label: fig-1
#| fig-cap: Street tree count of 10 most popoulus tree species in Toronto up to September 2024
#| echo: false
#| warning: false
#| message: false

# Summarize the model
summary(spline_model)

# Posterior predictive checks
pp_check(spline_model)
```

\newpage

# Discussion

## Trends from pollsters

## Trends from states

## Trends from sponsors

## Trends from methodology

## Weaknesses and next steps

\newpage

# Appendix

## Pollster Analysis

## Poll Blueprint

# References

